Notes

### 2022-09-17 chapters/bayesian-functional-regression.tex
Minor changes

\let\epsilon\varepsilon

@@

In this chapter we present the precise functional models and Bayesian methodologies explored in this work. The RKHS-based functional models under consideration \citep[see][]{berrendero2018functional, berrendero2019rkhs} are those obtained by substituting the functional parameter \(\beta\in L^2[0,1]\) for \(\alpha \in \Hcal(K)\), replacing also the scalar product \(\dotprod{X}{\beta}\) for \(\dotprod{X}{\alpha}_K\) in the \(L^2\)-models~\eqref{eq:l2-linear-model} and~\eqref{eq:l2-logistic-model}. However, to further simplify things we will follow a parametric approach and suppose that \(\alpha\) is in fact a member of the dense subspace \(\Hcal_0(K)\) defined in~\eqref{eq:h0}, i.e.:
\begin{equation}\label{eq:alpha_h0}
\alpha(\cdot) = \sum_{j=1}^p \beta_j K(t_j, \cdot), \text{ for some } p \in \N, \ \beta_j \in \R \text{ and } t_j \in [0,1].
\end{equation}
Moreover, as we said before, with a slight abuse of notation we will understand the expression \(\dotprod{x}{\alpha}_K\) as \(\Psi_x(\alpha)\), where \(x=X(\omega)\) and \(\Psi_x\) is Loève's isometry. Hence, taking into account that \(\Psi_X(K(t, \cdot)) = X(t)\) by definition (see~\eqref{eq:loeves-isometry}), we can write \(\dotprod{x}{\alpha}_K \equiv \sum_{j=1}^p \beta_j x(t_j)\) when \(\alpha\) is as in~\eqref{eq:alpha_h0}.

@@

At this point we propose to follow a Bayesian approach to estimate the parameters of the model, which we believe is in line with the idea of simplicity we pursue, and also introduces an additional layer of flexibility in the model. In this way, we can include problem-specific information in the model through the use of prior distributions, and on top of that, this method works almost unaltered for both linear and logistic regression models. The general idea will be to impose a prior distribution on the functional parameter to eventually derive a posterior model after incorporating the available sample information. In view of~\eqref{eq:alpha_h0}, to set a prior distribution on the unknown function \(\alpha\) (that is, a prior distribution on the functional space \(\Hcal_0(K)\)) it suffices to first consider a discrete distribution on \(p\), and then impose \(p\)-dimensional continuous prior distributions on the coefficients \(\beta_j\) and the times \(t_j\) given \(p\). Thanks to this parametric approach, the challenging task of setting a prior distribution on a space of functions is considerably simplified, while simultaneously not constraining the model to any specific distribution (in contrast to, for instance, Gaussian process regression methods). Moreover, note that starting from a probability distribution \(\P_0\) on \(\Hcal_0(K)\) we can obtain a probability distribution \(\P\) on \(\Hcal(K)\) merely by defining \(\P(B) = \P_0(B\cap \Hcal_0(K))\) for all Borel sets \(B\). Consequently, our simplifying assumption on \(\alpha\) is not very restrictive, since any prior distribution on \(\Hcal_0(K)\) can be directly extended to a prior distribution on \(\Hcal(K)\).

However, after some initial experimentation we found that, for practical and computational reasons, the value of \(p\), the dimensionality of the model, is best fixed beforehand in a suitable way (see Chapter~\ref{ch:model-choice} for details). Thus, we will regard only the \(\beta_j\) and \(t_j\) as free parameters, and search for our functional parameter in the space
\begin{equation}\label{eq:h0p}
\Hcal_{0,p}(K)=\left\{ \sum_{j=1}^p \beta_j K(t_j, \cdot): \ \beta_j \in \R, \ t_j \in [0, 1]\right\}.
\end{equation}
Even though we actually work on \(\Hcal_{0,p}(K)\), the discrete parameter \(p\) can still be selected in several meaningful ways that make use of the available data, and the set of feasible values is not very large in practice. Moreover, we could think of this approach as imposing a degenerate prior distribution on \(p\), so it is in a way a particular case of the more general model discussed above.

@@

In the case of functional logistic regression, we can regard the binary response variable \(Y\in\{0, 1\}\) as a Bernoulli random variable given \(X=x \in L^2[0, 1]\), and as usual suppose that \(\log\left(p(x)/(1-p(x))\right)\) is linear in \(x\), where \(p(x)=\P(Y=1| X=x)\). Then, following the approach  suggested by \citet{berrendero2018functional}, a RKHS model might be given, in terms of the correspondence \(\dotprod{X}{\alpha}_K = \Psi_X(\alpha)\), by the equation
\begin{equation}\label{eq:rkhs-model-logistic}
  \P(Y=1 \mid X) = \frac{1}{1 + \exp\{-\alpha_0 - \dotprod{X}{\alpha}_K\}}, \quad \alpha_0 \in \R, \ \alpha \in \Hcal_{0,p}(K).
\end{equation}

>>>>

\let\epsilon\varepsilon
\setlength{\footskip}{50.75pt}

@@

In this chapter we present the precise functional models and Bayesian methodologies explored in this work. The RKHS-based functional models under consideration \citep[see][]{berrendero2018functional, berrendero2019rkhs} are those obtained by substituting the functional parameter \(\beta\in L^2[0,1]\) for \(\alpha \in \Hcal(K)\), replacing also the scalar product \(\dotprod{X}{\beta}\) for \(\dotprod{X}{\alpha}_K\) in the \(L^2\)-models~\eqref{eq:l2-linear-model} and~\eqref{eq:l2-logistic-model}. However, to further simplify things we will follow a parametric approach and suppose that \(\alpha\) is in fact a member of the dense subspace \(\Hcal_0(K)\) defined in~\eqref{eq:h0}, i.e.:
\begin{equation}\label{eq:alpha_h0}
\alpha(\cdot) = \sum_{j=1}^p \beta_j K(t_j, \cdot), \text{ for some } p \in \N, \ \beta_j \in \R \text{ and } t_j \in [0,1].
\end{equation}
Moreover, as we said before, with a slight abuse of notation we will understand the expression \(\dotprod{x}{\alpha}_K\) as \(\Psi_x(\alpha)\), where \(x=X(\omega)\) and \(\Psi_x\) is Loève's isometry. Hence, taking into account that \(\Psi_X(K(t, \cdot)) = X(t)\) by definition (see~\eqref{eq:loeves-isometry}), when \(\alpha\) is as in~\eqref{eq:alpha_h0} we can write \(\dotprod{x}{\alpha}_K \equiv \sum_{j=1}^p \beta_j x(t_j)\).

@@

At this point we propose to follow a Bayesian approach to estimate the parameters of the model, which we believe is in line with the idea of simplicity we pursue, and also introduces an additional layer of flexibility in the model. In this way, we can include problem-specific information in the model through the use of prior distributions, and on top of that, this method works almost unaltered for both linear and logistic regression models. The general idea will be to impose a prior distribution on the functional parameter to eventually derive a posterior model after incorporating the available sample information.

In view of~\eqref{eq:alpha_h0}, to set a prior distribution on the unknown function \(\alpha\) (that is, a prior distribution on the functional space \(\Hcal_0(K)\)) it suffices to first consider a discrete distribution on \(p\), and then impose \(p\)-dimensional continuous prior distributions on the coefficients \(\beta_j\) and the times \(t_j\) given \(p\). Thanks to this parametric approach, the challenging task of setting a prior distribution on a space of functions is considerably simplified, while simultaneously not constraining the model to any specific distribution (in contrast to, for instance, Gaussian process regression methods). Moreover, note that starting from a probability distribution \(\P_0\) on \(\Hcal_0(K)\) we can obtain a probability distribution \(\P\) on \(\Hcal(K)\) merely by defining \(\P(B) = \P_0(B\cap \Hcal_0(K))\) for all Borel sets \(B\). Consequently, our simplifying assumption on \(\alpha\) is not very restrictive, since any prior distribution on \(\Hcal_0(K)\) can be directly extended to a prior distribution on \(\Hcal(K)\).

However, after some initial experimentation we found that, for practical and computational reasons, the value of \(p\) (the dimensionality of the model) is best fixed beforehand in a suitable way; see Chapter~\ref{ch:model-choice} for details. Thus, we will regard only the \(\beta_j\) and \(t_j\) as free parameters, and search for our functional parameter in the space
\begin{equation}\label{eq:h0p}
\Hcal_{0,p}(K)=\left\{ \sum_{j=1}^p \beta_j K(t_j, \cdot): \ \beta_j \in \R, \ t_j \in [0, 1]\right\}.
\end{equation}
Even though we actually work on \(\Hcal_{0,p}(K)\), the discrete parameter \(p\) can still be selected in several meaningful ways that make use of the available data, and the set of feasible values is not very large in practice. Moreover, we could think of this approach as imposing a degenerate prior distribution on \(p\), so it is in a way a particular case of the more general model discussed above.

@@

In the case of functional logistic regression, we regard the response variable \(Y\in\{0, 1\}\) as a Bernoulli random variable given \(X=x \in L^2[0, 1]\), and as usual suppose that \(\log\left(p(x)/(1-p(x))\right)\) is linear in \(x\), where \(p(x)=\P(Y=1| X=x)\). Then, following the approach  suggested by \citet{berrendero2018functional}, a RKHS model might be given, in terms of the correspondence \(\dotprod{X}{\alpha}_K = \Psi_X(\alpha)\), by the equation
\begin{equation}\label{eq:rkhs-model-logistic}
  \P(Y=1 \mid X) = \frac{1}{1 + \exp\{-\alpha_0 - \dotprod{X}{\alpha}_K\}}, \quad \alpha_0 \in \R, \ \alpha \in \Hcal_{0,p}(K).
\end{equation}

### 2022-09-17 chapters/conclusions.tex
Minor changes

In this work we have addressed some inference and prediction issues involved in the task of learning from functional data. In particular, we have tackled functional linear and logistic regression problems, and we have presented an alternative to the \(L^2\)-models predominant in these scenarios. Our proposed models rely heavily on the theory of RKHS's, offering finite-dimensional approximations but still with a functional perspective. This leads to a conceptually simpler approach that facilitates the interpretation and subsequent analysis, which we believe is a compelling feature for most practitioners. The underlying theoretical aspects are nevertheless far from trivial, so we have also explored some of the related mathematical background in this text.

@@

\begin{itemize}
  \item To delve further into the relationship between RKHS's and functional data problems, a connection that has proven to be fruitful in many scenarios. A first idea would be to extend the RKHS-based logistic regression model to a generalized functional linear model with an arbitrary link function.
  \item To try to derive some theoretical properties of our Bayesian predictors. For example, consistency and/or robustness results regarding the posterior distribution would be an excellent complement to the practical side of this work.
  \item To find other prior distributions for our parameters that perform better in general, or to eliminate the need of the tuning hyperparameter \(\eta\).
  \item To experiment with other MCMC algorithms for posterior approximation. Specifically, it would be very interesting to implement an efficient and reliable reversible-jump MCMC method in Python, which as we have already mentioned is a better fit for our particular Bayesian model.
\end{itemize}

>>>>

In this work we have addressed some inference and prediction issues involved in the task of learning from functional data. In particular, we have tackled functional linear and logistic regression problems, presenting an alternative to the \(L^2\)-models predominant in these scenarios. Our proposed models rely heavily on the theory of RKHS's, offering finite-dimensional approximations to the regression function, but still with a functional perspective. This leads to a conceptually simpler approach that facilitates the interpretation and subsequent analysis, which we believe is a compelling feature for most practitioners. The underlying theoretical aspects are nevertheless far from trivial, so we have also explored some of the related mathematical background in this text.

@@

\begin{itemize}
  \item To delve further into the relationship between RKHS's and functional data problems, a connection that has proven to be fruitful in many scenarios. A first idea would be to extend the RKHS-based logistic regression model to a generalized functional linear model with an arbitrary link function.
  \item To try to derive some theoretical properties of our Bayesian predictors. For example, consistency and/or robustness results regarding the posterior distribution would be an excellent complement to the practical side of this work.
  \item To find other prior distributions for our parameters that perform better in general, or to eliminate the need of the hyperparameters \(b_0\) or \(\eta\).
  \item To experiment with other MCMC algorithms for posterior approximation. Specifically, it would be very interesting to implement an efficient and reliable reversible-jump MCMC method in Python, which as we have already mentioned is a better fit for our particular Bayesian model.
\end{itemize}

### 2022-09-17 chapters/experiments.tex
Minor changes

\paragraph{Linear regression data sets.} We consider three different types of simulated data sets, all with a common value of \(\alpha_0=5\).
\begin{itemize}
  \item A RKHS model where the response is generated as \(Y= 0.5 -5X(0.1) + X(0.4) + 10X(0.8) + \epsilon\), for each of the four GP's mentioned above.
  \item A \(L^2\)-model with underlying coefficient function \(\beta(t)=\log(1+4t)\), again for the same four GP's.
  \item A model based on non-GP regressors. Specifically, we use a geometric Brownian motion (GBM) defined as \(X(t)=e^{BM(t)}\), where \(BM(t)\) is a standard Brownian motion. In this case, we consider two data sets, one with a RKHS response and one with a \(L^2\) response, with the same parameters as above.
\end{itemize}
As for the real data sets, we use the (twice-differentiated) Tecator data set introduced in Chapter~\ref{ch:introduction} to predict fat content of 193 meat samples, as well as what we call the Moisture \citep{kalivas1997two} and Sugar \citep{bro1999exploratory} data sets. The first consists of near-infrared spectra of 100 wheat samples and the objective is to predict the samples' moisture content, whereas the second contains 268 samples of sugar fluorescence data in order to predict ash content. The three data sets are measured on a grid of 100, 101 and 115 equispaced points on \([0, 1]\), respectively.

@@

Take for example a RKHS data set with two components generated according to the formula \(Y=-5X(0.1) + 10X(0.8) + \epsilon\), with \(\epsilon\sim \mathcal N(0, 0.5)\). Figure~\ref{fig:beta_trace_3} shows the resulting posterior distribution of the parameters \(b=(\beta_1, \beta_2, \beta_3)\) and \(\tau=(\tau_1, \tau_2, \tau_3)\) for a model with 3 components. As we can see, one of the coefficients has gone to zero to account for the overspecification of the model, while the other two have stabilized very close to the true parameters. The same goes for the time instants, except that in this case there is no default value to represent that a component is unused, so the time corresponding to the null coefficient oscillates back and forth. The estimated function (based for example in the mode of the posterior distributions) will not be perfect, essentially because of the noise in the response. But it should be close to the true parameter function \(\alpha(t)=-5K(t, 0.1) + 10K(t, 0.8)\), providing a good predictive performance.

>>>>

\paragraph{Linear regression data sets.} We consider three different types of simulated data sets, all with a common value of \(\alpha_0=5\).
\begin{itemize}
  \item A RKHS model where the response is generated as \(Y=5 -5X(0.1) + X(0.4) + 10X(0.8) + \epsilon\), for each of the four GP's mentioned above.
  \item A \(L^2\)-model with underlying coefficient function \(\beta(t)=\log(1+4t)\), again for the same four GP's.
  \item A model based on non-GP regressors. Specifically, we use a geometric Brownian motion (GBM) defined as \(X(t)=e^{BM(t)}\), where \(BM(t)\) is a standard Brownian motion. In this case, we consider two data sets, one with a RKHS response and one with a \(L^2\) response, with the same parameters as above.
\end{itemize}
As for the real data sets, we use the (twice-differentiated) Tecator data set introduced in Chapter~\ref{ch:introduction} to predict fat content of 193 meat samples, as well as what we call the Moisture \citep{kalivas1997two} and Sugar \citep{bro1999exploratory} data sets. The first consists of near-infrared spectra of 100 wheat samples and the objective is to predict the samples' moisture content, whereas the second contains 268 samples of sugar fluorescence data in order to predict ash content. The three data sets are measured on a grid of 100, 101 and 115 equispaced points on \([0, 1]\), respectively.

@@

Take for example a RKHS data set with two components generated according to the formula \(Y=5 -5X(0.1) + 10X(0.8) + \epsilon\), with \(\epsilon\sim \mathcal N(0, 0.5)\). Figure~\ref{fig:beta_trace_3} shows the resulting posterior distribution of the parameters \(b=(\beta_1, \beta_2, \beta_3)\) and \(\tau=(\tau_1, \tau_2, \tau_3)\) for a model with 3 components. As we can see, one of the coefficients has gone to zero to account for the overspecification of the model, while the other two have stabilized very close to the true parameters. The same goes for the time instants, except that in this case there is no default value to represent that a component is unused, so the time corresponding to the null coefficient oscillates back and forth. The estimated function (based for example in the mode of the posterior distributions) will not be perfect, essentially because of the noise in the response. But it should be close to the true parameter function \(\alpha(t)=-5K(t, 0.1) + 10K(t, 0.8)\), providing a good predictive performance.
